Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2021-01-26 xrb2/paper.tex
minor changes

We investigate the effect of rotation rate on the flame. With the
exception of IGR J17480-2446 (\citealt{altamirano2010discovery}, spinning at $11~\mathrm{Hz}$), most observations of
XRBs come from sources with rotation rates of $200-600~\mathrm{Hz}$
\citep{bilous:2019,galloway:2020}. There are a number of
factors that could explain this lack of observations below $200~\mathrm{Hz}$. It
could be that there is some physical process which inhibits the flame
ignition and/or spread at lower rotation rates. It could be that burst
oscillations at lower rotation rates are smaller in amplitude and
therefore more difficult to detect. It could be that it does not have
anything to do with the flame at all, but that neutron stars in
accreting low mass X-ray binaries rarely have rotation rates below $200~\mathrm{Hz}$.

@@

We use the Castro hydrodynamics code \citep{castro,castro_joss} and
the simulation framework introduced in \citet{flame_wave1}.  The
current simulations are all performed in a two-dimensional
axisymmetric geometry.  For these axisymmetric simulations, we add an
additional geometric source term from \citet{bernard-champmartin} that
captures the effects of the divergence of the flux operating on the
azimuthal unit vector.  This term is a small correction, but was
missing from our previous simulations.  The simulation framework
initializes a fuel layer in hydrostatic equilibrium, laterally
blending a hot model on the left side of the domain (the coordinate
origin) and a cool model on the right.  The initial temperature
gradient between the hot and cool fluids drives a laterally
propagating flame through the cool fuel.  In our original set of
calculations (in \citet{flame_wave1}), in order to make the 
simulations computationally feasible we artificially boosted the flame speed by adjusting the
conductivity and reaction rate to produce a flame moving 5--10$\times$
faster than the nominal laminar flame speed.  We also used high
rotation rates ($\geq 1000~\mathrm{Hz}$) to reduce the lateral lengthscale at which the Coriolis
force balances the lateral flame spreading in order to reduce the size 
of the simulation domain. The port of \castro\ to GPUs~\citep{sc20_gpu} 
significantly improved its overall performance, enabling
us to run these new simulations without the previous approximations while
continuing to resolve the burning front. For these
simulations, we no longer boost the flame speed---the true
conductivities (taken from \citealt{Timmes00}) and reaction rates are
used. We are also able to use slower, more physically realistic rotation rates.
  We continue to use a 13-isotope $\alpha$-chain to describe the
helium burning.

@@

We present six simulations in total, summarized in
Table~\ref{table:runs}.  These simulations encompass three different
rotation rates: $250~\mathrm{Hz}$, $500~\mathrm{Hz}$, and $1000~\mathrm{Hz}$, 
and for the $500~\mathrm{Hz}$ run, four
different temperature profiles.  In the following subsections, we look
at how the flame properties depend on the model parameters.  All
simulations are run in a domain of $1.8432\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$ with a coarse grid of $1152 \times 192$
zones and two levels of refinement (the first level refining the resolution 
by factor of four, and the second by a factor of two again).  This gives a
fine-grid resolution of $20~\mathrm{cm}$.  In these simulations, refinement 
is carried out in all zones within the atmosphere
with $\rho > 2.5\times 10^4~\gcc$. We use an axisymmetric coordinate system, 
with the horizontal $r$-direction pointing along the surface of the star and the 
vertical $z$-direction pointing tangential to the surface.

@@

For some of our analysis, we would like to have a means of estimating the temperature ($T$) and 
nuclear energy generation rate ($\enucdot$) in the burning region of each simulation. For this 
purpose, we define the mass-weighted average $\repavg{Q}$ of some quantity $Q$ to be
\begin{equation}
	\label{eqn:top_percentile_avg}
	\repavg{Q} \equiv \frac{\sum_{c_i} m(c_i) Q(c_i)}{\sum_{c_i} m(c_i)}; ~c_i \in C_{99}(Q).
\end{equation}
\noindent Here, $C_{99}(Q)$ is the set of grid cells with $Q$ values in the top percentile, and 
$Q(c_i)$ is the value of $Q$ in cell $c_i$ (with $m(c_i)$ being the total mass in cell $c_i$). Using $\repavg{Q}$ instead of simply taking the maximum of the quantity across the entire 
simulation domain allows us to track changes over the domain as a whole rather than at a single 
localized point. This will therefore be a better reflection of the overall behavior of the flame 
rather than of a single localized fluctuation. Figure \ref{fig:maxima} shows $\repavg{T}$ and 
$\repavg{\enucdot}$ as functions of time for the subset of our runs that achieve a propagating 
flame. This figure is referenced throughout the subsequent sections.

@@

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. We saw in \citet{flame_wave1} that 
increasing the neutron star rotation rate reduces the horizontal lengthscale of the flame. An 
estimate of this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius 
may be thought of as the scale over which the balance between the Coriolis force and horizontal 
pressure gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is the gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use the $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region where the most significant energy generation is occurring as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). From Equation 
\ref{eqn:rossby}, we can see that increasing the rotation rate from $500~\mathrm{Hz}$ to 
$1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater confinement from 
the Coriolis force should reduce the horizontal extent of the flame by a similar factor. However, 
the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our 
simulations we see that this scaling does not work so well for all rotation rates. The simulations 
seem to follow the theoretical scaling described in Equation \ref{eqn:rossby} more closely at 
higher rotation rates ($1000~\mathrm{Hz}$ and higher), based on the results of \citet{flame_wave1}.

@@

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} $100~\mathrm{ms}$ snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. The $\dot{e}_\mathrm{nuc}$ plots in Figure 
\ref{fig:time_series_enuc_1000} also support this interpretation, with the region of the 
flame front nearest to the crust in the {\tt F1000} run reaching higher $\dot{e}_\mathrm{nuc}$ values 
than for the {\tt F500} run in Figure~\ref{fig:time_series_enuc_500}. In contrast to {\tt F500} 
and {\tt F1000}, the lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition 
for {\tt F250} also aligns with the reasoning given above, with the lower rotation in this case 
potentially leading to insufficient confinement such that the temperature and density required for 
ignition is not achieved. In this scenario, another source of confinement (e.g.\ magnetic fields, see \citep{art-2016-cavecchi-etal}) 
would need to take over at lower rotation rates to allow a burst to occur, at least for the initial 
model used here. Given that the size of our domain is $\sim 1~L_R$ for {\tt F250} (using Equation~\ref{eqn:rossby}), it is also possible that we simply cannot confine the flame sufficiently with our current domain width. We see in Figure~\ref{fig:flame_speeds_1} (discussed further in Section~\ref{ssec:rot_propagation}) that the {\tt F500} flame took longer to achieve steady propagation than the {\tt F1000} flame. It may therefore also be that we did not run our simulation for long enough to see the {\tt F250} flame achieve the conditions required for ignition and steady propagation.

Burning in the {\tt F500} and {\tt F1000} runs is concentrated in a dense region with circular 
motion. In Figure~\ref{fig:urho}, which compares the horizontal $r$-velocity $u$, density $\rho$ and the 
nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the {\tt F250}, {\tt F500}, and {\tt F1000} 
runs, most of the burning for each of the simulations occurs in a high density region $\rho > 3 
\times 10^5 {\rm g\,cm^{-3}}$. The fluid in this dense, high energy generation region undergoes 
vortical motion, shown in the Figure~\ref{fig:uv} phase plots comparing $u$, the vertical $z$-velocity $v$ and $\dot{e}_{\rm 
nuc}$. This most likely corresponds to the leading edge of the flame where fresh fuel is being entrained.
This feature is not developed in the $250~\mathrm{Hz}$ flame in Figure~\ref{fig:uv} (left panel); it could potentially develop 
at later times (past the point at which we terminated our simulation), or the flame could just 
fizzle out and fail to ignite entirely.

>>>>

We investigate the effect of rotation rate on the flame. With the
exception of IGR J17480-2446 (\citealt{altamirano2010discovery}, spinning at $11~\mathrm{Hz}$), most observations of
XRBs come from sources with rotation rates of $200-600~\mathrm{Hz}$
\citep{bilous:2019,galloway:2020}. There are a number of
factors that could explain this lack of observations below $200~\mathrm{Hz}$. It
could be that there is some physical process which inhibits the flame
ignition and/or spread at lower rotation rates. It could be that bursts
at lower rotation rates are smaller in amplitude and
therefore more difficult to detect. It could be that it does not have
anything to do with the flame at all, but that neutron stars in
accreting low mass X-ray binaries rarely have rotation rates below $200~\mathrm{Hz}$.

@@

We use the Castro hydrodynamics code \citep{castro,castro_joss} and
the simulation framework introduced in \citet{flame_wave1}.  The
current simulations are all performed in a two-dimensional
axisymmetric geometry.  For these axisymmetric simulations, we add an
additional geometric source term from \citet{bernard-champmartin} that
captures the effects of the divergence of the flux operating on the
azimuthal unit vector.  This term is a small correction, but was
missing from our previous simulations.  The simulation framework
initializes a fuel layer in hydrostatic equilibrium, laterally
blending a hot model on the left side of the domain (the coordinate
origin) and a cool model on the right.  The initial temperature
gradient between the hot and cool fluids drives a laterally
propagating flame through the cool fuel.  In our original set of
calculations (in \citep{flame_wave1}), in order to make the 
simulations computationally feasible we artificially boosted the flame speed by adjusting the
conductivity and reaction rate to produce a flame moving 5--10$\times$
faster than the nominal laminar flame speed.  We also used high
rotation rates ($\geq 1000~\mathrm{Hz}$) to reduce the lateral lengthscale at which the Coriolis
force balances the lateral flame spreading in order to reduce the size 
of the simulation domain. The port of \castro\ to GPUs~\citep{sc20_gpu} 
significantly improved its overall performance, enabling
us to run these new simulations without the previous approximations while
continuing to resolve the burning front. For these
simulations, we no longer boost the flame speed---the true
conductivities (taken from \citealt{Timmes00}) and reaction rates are
used. We are also able to use slower, more physically realistic rotation rates.
  We continue to use a 13-isotope $\alpha$-chain to describe the
helium burning.

@@

We present six simulations in total, summarized in
Table~\ref{table:runs}.  These simulations encompass three different
rotation rates: $250~\mathrm{Hz}$, $500~\mathrm{Hz}$, and $1000~\mathrm{Hz}$, 
and for the $500~\mathrm{Hz}$ run, four
different temperature profiles.  In the following subsections, we look
at how the flame properties depend on the model parameters.  All
simulations are run in a domain of $1.8432\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$ with a coarse grid of $1152 \times 192$
zones and two levels of refinement (the first level refining the resolution 
by factor of four, and the second by a factor of two again).  This gives a
fine-grid resolution of $20~\mathrm{cm}$.  In these simulations, refinement 
is carried out in all zones within the atmosphere
with density $\rho > 2.5\times 10^4~\gcc$. We use an axisymmetric coordinate system, 
with the horizontal $r$-direction pointing along the surface of the star and the 
vertical $z$-direction pointing tangential to the surface.

@@

For some of our analysis, we would like to have a means of estimating the temperature ($T$) and 
nuclear energy generation rate ($\enucdot$) in the burning region of each simulation. For this 
purpose, we define the mass-weighted average $\repavg{Q}$ of some quantity $Q$ to be
\begin{equation}
	\label{eqn:top_percentile_avg}
	\repavg{Q} \equiv \frac{\sum_{c_i} m(c_i) Q(c_i)}{\sum_{c_i} m(c_i)}; ~c_i \in C_{99}(Q).
\end{equation}
\noindent Here, $C_{99}(Q)$ is the set of grid cells with $Q$ values in the top percentile, 
$Q(c_i)$ is the value of $Q$ in cell $c_i$, and $m(c_i)$ is the total mass in cell $c_i$. 
Using $\repavg{Q}$ instead of simply taking the maximum of the quantity across the entire 
simulation domain allows us to track changes over the domain as a whole rather than at a single 
localized point. This will therefore be a better reflection of the overall behavior of the flame 
rather than of a single localized fluctuation. Figure \ref{fig:maxima} shows $\repavg{T}$ and 
$\repavg{\enucdot}$ as functions of time for the subset of our runs that achieve a propagating 
flame. This figure is referenced throughout the subsequent sections.

@@

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. We saw in \citet{flame_wave1} that 
increasing the neutron star rotation rate reduces the horizontal lengthscale of the flame. An 
estimate of this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius 
may be thought of as the scale over which the balance between the Coriolis force and horizontal 
pressure gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is the gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region where the most significant energy generation is occurring as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). From Equation 
\ref{eqn:rossby}, we can see that increasing the rotation rate from $500~\mathrm{Hz}$ to 
$1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater confinement from 
the Coriolis force should reduce the horizontal extent of the flame by a similar factor. However, 
the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our 
simulations we see that this scaling does not work so well for all rotation rates. The simulations 
seem to follow the theoretical scaling described in Equation \ref{eqn:rossby} more closely at 
higher rotation rates ($1000~\mathrm{Hz}$ and higher), based on the results of \citet{flame_wave1}.

@@

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} $100~\mathrm{ms}$ snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. The $\dot{e}_\mathrm{nuc}$ plots in Figure 
\ref{fig:time_series_enuc_1000} also support this interpretation, with the region of the 
flame front nearest to the crust in the {\tt F1000} run reaching higher $\dot{e}_\mathrm{nuc}$ values 
than for the {\tt F500} run in Figure~\ref{fig:time_series_enuc_500}. In contrast to {\tt F500} 
and {\tt F1000}, the lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition 
for {\tt F250} also aligns with the reasoning given above, with the lower rotation in this case 
potentially leading to insufficient confinement such that the temperature and density required for 
ignition is not achieved. In this scenario, another source of confinement (e.g.\ magnetic fields, see \cite{art-2016-cavecchi-etal}) 
would need to take over at lower rotation rates to allow a burst to occur, at least for the initial 
model used here. Given that the size of our domain is $\sim 1~L_R$ for {\tt F250} (using Equation~\ref{eqn:rossby}), it is also possible that we simply cannot confine the flame sufficiently with our current domain width. We see in Figure~\ref{fig:flame_speeds_1} (discussed further in Section~\ref{ssec:rot_propagation}) that the {\tt F500} flame took longer to achieve steady propagation than the {\tt F1000} flame. It may therefore also be that we did not run our simulation for long enough to see the {\tt F250} flame achieve the conditions required for ignition and steady propagation.

Burning in the {\tt F500} and {\tt F1000} runs is concentrated in a dense region with circular 
motion. In Figure~\ref{fig:urho}, which compares the horizontal $r$-velocity $u$, density $\rho$ and the 
nuclear energy generation rate $\dot{e}_{\rm nuc}$ for the {\tt F250}, {\tt F500}, and {\tt F1000} 
runs, most of the burning for each of the simulations occurs in a high density region $\rho > 3 
\times 10^5 {\rm g\,cm^{-3}}$. The fluid in this dense, high energy generation region undergoes 
vortical motion, shown in the Figure~\ref{fig:uv} phase plots comparing $u$, the vertical $z$-velocity $v$ and $\dot{e}_{\rm 
nuc}$. This most likely corresponds to the leading edge of the flame where fresh fuel is being entrained.
This feature is not developed in the $250~\mathrm{Hz}$ flame in Figure~\ref{fig:uv} (left panel); it could potentially develop 
at later times (past the point at which we terminated our simulation), or the burning could just 
fizzle out and fail to ignite entirely.
